Selective Sequential Model Selection

### 2015-10-19 paper.tex
rob edited sec 1-5

The max-$t$ test $p$-values are computed using the same test statistic $|t_E^*|$, but compared with a more appropriate null distribution. We can simulate the distribution of $|t_E^*|$ under the null model, i.e. the linear regression model specified by active set $E$:
\begin{equation}
  M(E):\; Y \sim N(X_E\beta, \sigma^2 I), \quad \beta\in \R^{|E|}, \sigma^2>0,
\end{equation}
where $X_E$ is the matrix with columns $(X_j)_{j\in E}$. Because $U_E=\left(X_E'Y,\, \|Y\|^2\right)$ is the complete sufficient statistic for $M(E)$, we can sample from the conditional null distribution of $t_E^*$ given $U_E$ and the current active set $E$, using the {\em selected-model test} framework of \citet{fithian2014optimal}. In step one, because $E=\{1_n\}$ is fixed and $t_E^*$ is independent of $U_E$, the conditional test reduces to the max-$t$ test proposed in \citet{buja2014}. In later steps, $t_E^*$ and $U_E$ are not conditionally independent of $E$, but we can determine the null conditional distribution through Monte Carlo sampling.

@@

In comparison to their saturated-model counterparts, we will see that the selected-model tests proposed here have three main advantages: power, independence (in some cases), and generalizability beyond linear regression. These advantages do not come entirely for free, as we will see: the selected-model methods test a more restrictive null hypothesis than the saturated-model methods.

@@

In the general setting, we observe data $Y \in \cY$ with unknown sampling distribution $F$, and then apply some algorithm to generate an adaptive sequence of $d$ nested statistical models contained in some {\em upper model} $M_{\infty}$.
\[
M_0(Y) \sub M_1(Y) \sub \cdots \sub M_d(Y) \sub M_\infty.
\]
By {\em model}, we mean a family of candidate probability distributions for $Y$. Without loss of generality, we can take $M_\infty$ as the union of all models under consideration. We will use the notation $M_{0:d}$ to denote the sequence $M_0, \ldots, M_d$, which we call the {\em model path}. We assume throughout that for each candidate model $M$ there is a minimal sufficient statistic $U(Y; \,M)$, and write
\[
U_k(Y) = U(Y; \,M_k(Y)).
\]

@@

\subsubsection{Ever-Active Regularization Paths and next-$\lambda$ Tests}
Another important class of model selection procedures is the sequence of {\em ever-active} sets for a regularized likelihood path. For some ordered set $\Lambda$, let $\lambda\in\Lambda$ parameterize the regularization penalty $P_\lambda(\theta)$. In the case of lasso-penalized regression, we could take $P_\lambda = \lambda\|\beta\|_1$ for $\lambda\in (0,\infty]$.

We then define
\begin{align}
  \hat\theta^{\lambda}(Y) &= 
  \argmin_{\theta\in\Theta} -\ell(\theta; Y) + P_\lambda(\theta) \\
  \tE_\lambda(Y) &= \left\{j:\; \hat\theta_j^\nu \neq 0 
    \text{ for any } \nu\in \Lambda, \nu \geq \lambda \right\}
\end{align}
It may be impossible or inconvenient to compute $\hat\theta^\lambda$ for every $\lambda\in(0,\infty]$. If so, we can instead take $\Lambda$ to be a grid of finitely many positive values.

@@

\citet{gsell2013sequential} describe three different null hypotheses that we could consider testing at step $k$ in the case of linear regression:
\begin{description}
\item[Complete Null:] $M_{k-1}$ is already correct. That is, 
\[
H_k:\;\mu = X_{E_{k-1}} \theta^{E_{k-1}}.
\]
\item[Incremental Null:] $M_{k-1}$ may be incorrect, but $M_k$ is no improvement. That is, 
\[
H_k^{\text{inc}}:\; \theta_{j_k}^{E_k} = 0.
\]
\item[Full-Model Null:] The coefficient of $X_{j_k}$ is zero in the ``full'' model with all $p$ predictors. That is,
\[
H_k^{\text{full}}:\; \theta_{j_k}^{\{1,\ldots,p\}} = 0.
\]
\end{description}

@@

In the second procedure, we fit the full model $M(\{1,\ldots,p\})$ and compute a $t$-statistic for each variable with respect to that model, yielding $t$-statistics $t_1, t_2, \ldots, t_p$ (not ordered by size). Iterating for $j=1,\ldots,p$, we compare each $|t_j|$ to the level-$\alpha$ cutoff for a $t$-test with $n-p$ degrees of freedom, and stop selecting variables the first time $|t_{k+1}|$ is not larger than this cutoff. This procedure would control the full-model-null FWER, but {\em not} the incremental-null FWER. Note that the {\em model} $M_k$ plays no role in this procedure: its fit to the data is never evaluated and it is not really selected in any meaningful sense. Even if the fit of $M_k$ is far worse than that of $M_{k+1}$, we may not be able to select $X_{k+1}$ if (say) it is highly collinear with $X_{k+100}$. Conversely, even if the fit of $M_k$ is no better than that of $M_{k-1}$, we may still select $X_k$ if it is a highly useful predictor when used in concert with $X_{k+100}$.

More generally, methods that control full-model error rates are best viewed not as {\em model-selection} procedures --- since all inferences are made with respect to the full model, which is specified ahead of time --- but {\em variable-selection} procedures that test multiple hypotheses with respect to a single fixed model. Note that the truth or falsity of these hypotheses can depend sensitively on the set of $p$ predictors: rejecting $H_j^{\text{full}}$ has no meaning without reference to the list of all variables that we controlled for, which can make a rejection difficult to interpret when $p$ is large. Thus, error rates like $\text{FDR}^{\text{full}}$ are best motivated when the full model has some special scientific status. For example, the scientist may believe due to theoretical considerations that the linear model in $X_1,\ldots,X_p$ is fairly credible, and that a nonzero coefficient $\beta_j$, controlling for all of the other variables, would constitute evidence for a causal effect of $X_j$ on the response.

@@

The problem of model selection is an old one, with quite an extensive literature. However, with the exception of the works above, very few methods offer finite-sample guarantees on the model that is selected except in the orthogonal-design case. One additional exception is the knockoff filter of \citet{barber2014controlling}, a variant of which provably controls the full-model FDR.

Methods like AIC \citet{akaike1974new} and BIC \citep{schwarz1978estimating} are designed for the non-adaptive case, where the sequence of models is determined in advance of observing the data. Cross-validation, another general-purpose algorithm for tuning parameter selection, targets out-of-sample error and tends to select many noise variables when the signal is sparse \WFcomment{Whom to cite?}. \citet{benjamini2009simple} extend the AIC by using an adaptive penalty $\sigma^2\lambda_{k,p} k$ to select a model, based on generalizing the Benjamini-Hochberg procedure \citet{benjamini1995controlling}, but do not prove finite-sample control. The stability selection approach of \citet{meinshausen2010stability} uses an approach based on splitting the data many times and offers asymptotic control of the FDR, but no finite-sample guarantees are available.

@@

In our problem setting, the sequence of selected models is random; thus, the truth or falsehood of each $H_k$ is not fixed, as assumed in the previous section. This could in principle lead to violation of the error control guarantees that hold in the fixed-hypothesis case.

To see why we a problem could arise, consider apply BasicStop in the following counterexample: let $d=2$ and let $k_0$ be 0 or 1 with probability 1/2 each. Then, let 
\[
p_1 \sim U[0,1], \quad p_2 \sim (1-k_0 + U[0,1])/2.
\]
Both $p_1$ and $p_2$ are uniform, but
\begin{align*}
\P(V>0) &= \P(p_{k_0+1} \leq \alpha)\\ 
&= \sum_{k=1}^2\P(p_k\leq \alpha \mid k_0+1=k)\cdot \P(k_0+1=k)\\
&= \alpha\cdot \frac{1}{2} + 2\alpha \cdot \frac{1}{2} = 3\alpha/2
\end{align*}
This problem is resolved if $p_k$ is a valid $p$-value conditional on the value of $k_0$.

@@

The sufficient filtration separates $p_{1:d}$ if and only if 
\begin{enumerate}
\item $p_k$ is conservative given $M_{0:k-1}$ and $U_{k-1}$, and
\item $p_k$ is a function of $M_{0:k}$ and $U_k$.
\end{enumerate}

@@

\section{Discussion}\label{sec:discussion}

>>>>

The max-$t$ test $p$-values are computed using the same test statistic $|t_E^*|$, but compared with a more appropriate null distribution. We can simulate the distribution of $|t_E^*|$ under the null model, i.e. the linear regression model specified by active set $E$:
\begin{equation}
  M(E):\; Y \sim N(X_E\beta, \sigma^2 I), \quad \beta\in \R^{|E|}, \sigma^2>0,
\end{equation}
where $X_E$ is the matrix with columns $(X_j)_{j\in E}$. Because $U_E=\left(X_E'Y,\, \|Y\|^2\right)$ is the complete sufficient statistic for $M(E)$, we can sample from the conditional null distribution of $t_E^*$ given $U_E$ and the current active set $E$, using the {\em selected-model test} framework of \citet{fithian2014optimal}. In step one, because $E=\{1_n\}$ is fixed and $t_E^*$ is independent of $U_E$, the conditional test reduces to the max-$t$ test proposed in \citet{buja2014}. In later steps, $t_E^*$ and $U_E$ are not conditionally independent of $E$, but we can estimate the null conditional distribution through Monte Carlo sampling.

@@

In comparison to their saturated-model counterparts, we will see that the selected-model tests proposed here have three main advantages: power, independence (in some cases), and generalizability beyond linear regression. These advantages do not come entirely for free, as we will see: the selected-model methods test a more restrictive null hypothesis than the saturated-model methods. In addition, the selected-model test require accept-reject or MCMC sampling, while the saturated-model tests are available in closed form.

@@

In the general setting, we observe data $Y \in \cY$ with unknown sampling distribution $F$, and then apply some algorithm to generate an adaptive sequence of $d$ nested statistical models contained in some {\em upper model} $M_{\infty}$.
\[
M_0(Y) \sub M_1(Y) \sub \cdots \sub M_d(Y) \sub M_\infty.
\]
By {\em model}, we mean a family of candidate probability distributions for $Y$. For example in  linear regression, a model specifies  the set of
predictors with nonzero coefficients (but not the  values of their coefficients). Without loss of generality, we can take $M_\infty$ as the union of all models under consideration. We will use the notation $M_{0:d}$ to denote the sequence $M_0, \ldots, M_d$, which we call the {\em model path}. We assume throughout that for each candidate model $M$ there is a minimal sufficient statistic $U(Y; \,M)$, and write
\[
U_k(Y) = U(Y; \,M_k(Y)).
\]

@@

\subsubsection{Ever-Active Regularization Paths and next-$\lambda$ Tests}
Another important class of model selection procedures is the sequence of {\em ever-active} sets for a regularized likelihood path. 
The notion of ever-active sets is needed since these solution paths can drop  (as well as add) predictors along the way. 	For some ordered set $\Lambda$, let $\lambda\in\Lambda$ parameterize the regularization penalty $P_\lambda(\theta)$. In the case of lasso-penalized regression, we could take $P_\lambda = \lambda\|\beta\|_1$ for $\lambda\in (0,\infty]$. 
We assume $P_\lambda(\beta)$  has the property  that the model becomes less constrained as $\lambda$ decreases, that is,
 if $\lambda_2<\lambda_1$ then $\{\beta:P_{\lambda_2}(\beta)\} \supset \{\beta:P_{\lambda_1}(\beta)\}$.  \RTcomment{Will: is this correct?}

We then define
\begin{align}
  \hat\theta^{\lambda}(Y) &= 
  \argmin_{\theta\in\Theta} -\ell(\theta; Y) + P_\lambda(\theta) \\
  \tE_\lambda(Y) &= \left\{j:\; \hat\theta_j^\nu(Y) \neq 0 
    \text{ for any } \nu\in \Lambda, \nu \geq \lambda \right\}
\end{align}
It may be impossible or inconvenient to compute $\hat\theta^\lambda(Y)$ for every $\lambda\in(0,\infty]$. If so, we can instead take $\Lambda$ to be a grid of finitely many positive values.

@@

\citet{gsell2013sequential} describe three different null hypotheses that we could consider testing at step $k$ in the case of linear regression:
\begin{description}
\item[Complete Null:] $M_{k-1}$ is (already) correct. That is, 
\[
H_k:\;\mu = X_{E_{k-1}} \theta^{E_{k-1}}.
\]
\item[Incremental Null:] $M_{k-1}$ may be incorrect, but $M_k$ is no improvement. That is, 
\[
H_k^{\text{inc}}:\; \theta_{j_k}^{E_k} = 0.
\]
\item[Full-Model Null:] The coefficient of $X_{j_k}$ is zero in the ``full'' model with all $p$ predictors. That is,
\[
H_k^{\text{full}}:\; \theta_{j_k}^{\{1,\ldots,p\}} = 0.
\]
\end{description}

@@

In the second procedure, we fit the full model $M(\{1,\ldots,p\})$ and compute a $t$-statistic for each variable with respect to that model, yielding $t$-statistics $t_1, t_2, \ldots, t_p$ (not ordered by size). For $k=1,\ldots,p$, we compare each $|t_j|$ to the level-$\alpha$ cutoff for a $t$-test with $n-p$ degrees of freedom, and stop selecting variables the first time $|t_{k+1}|$ is not larger than this cutoff. We might use this procedure if the variables had a  fixed ordering.
This procedure would control the full-model-null FWER, but {\em not} the incremental-null FWER. Note that the {\em model} $M_k$ plays no role in this procedure: its fit to the data is never evaluated and it is not really selected in any meaningful sense. Even if the fit of $M_k$ is far worse than that of $M_{k+1}$, we may not be able to select $X_{k+1}$ if (say) it is highly collinear with $X_{k+100}$. Conversely, even if the fit of $M_k$ is no better than that of $M_{k-1}$, we may still select $X_k$ if it is a highly useful predictor when used in concert with $X_{k+100}$.

More generally, methods that control full-model error rates are best viewed not as {\em model-selection} procedures --- since all inferences are made with respect to the full model, which is specified ahead of time --- but {\em variable-selection} procedures that test multiple hypotheses with respect to a single fixed model. Note that the truth or falsehood of these hypotheses can depend sensitively on the set of $p$ predictors: rejecting $H_j^{\text{full}}$ has no meaning without reference to the list of all variables that we controlled for, which can make a rejection difficult to interpret when $p$ is large. Thus, error rates like $\text{FDR}^{\text{full}}$ are best motivated when the full model has some special scientific status. For example, the scientist may believe due to theoretical considerations that the linear model in $X_1,\ldots,X_p$ is fairly credible, and that a nonzero coefficient $\beta_j$, controlling for all of the other variables, would constitute evidence for a causal effect of $X_j$ on the response.

@@

The problem of model selection is an old one, with quite an extensive literature. However, with the exception of the works above, very few methods offer finite-sample guarantees on the model that is selected except in the orthogonal-design case. One  exception is the knockoff filter of \citet{barber2014controlling}, a variant of which provably controls the full-model FDR.
We compare our proposal to the Knockoff method in Section \ref{sec:sparseReg}.

Methods like AIC \citet{akaike1974new} and BIC \citep{schwarz1978estimating} are designed for the non-adaptive case, where the sequence of models is determined in advance of observing the data. Cross-validation, another general-purpose algorithm for tuning parameter selection, targets out-of-sample error and tends to select many noise variables when the signal is sparse (e.g. \cite{LY2015}).   \citet{benjamini2009simple} extend the AIC by using an adaptive penalty $\sigma^2\lambda_{k,p} k$ to select a model, based on generalizing the Benjamini-Hochberg procedure \citet{benjamini1995controlling}, but do not prove finite-sample control. The stability selection approach of \citet{meinshausen2010stability} uses an approach based on splitting the data many times and offers asymptotic control of the FDR, but no finite-sample guarantees are available.

@@

In our problem setting, the sequence of selected models is random; thus, the truth or falsehood of each $H_k$ is not fixed, as assumed in the previous section. This could in principle lead to violation of the error control guarantees that hold in the fixed-hypothesis case.
\RTcomment{As Will and I discussed, this example is confusing and needs improvement.}
To see why we a problem could arise, consider apply BasicStop in the following counterexample: let $d=2$ and let $k_0$ be 0 or 1 with probability 1/2 each. Then, let 
\[
p_1 \sim U[0,1], \quad p_2 \sim (1-k_0 + U[0,1])/2.
\]
Both $p_1$ and $p_2$ are uniform, but
\begin{align*}
\P(V>0) &= \P(p_{k_0+1} \leq \alpha)\\ 
&= \sum_{k=1}^2\P(p_k\leq \alpha \mid k_0+1=k)\cdot \P(k_0+1=k)\\
&= \alpha\cdot \frac{1}{2} + 2\alpha \cdot \frac{1}{2} = 3\alpha/2
\end{align*}
This problem is resolved if $p_k$ is a valid $p$-value conditional on the value of $k_0$.

@@

The sufficient filtration separates $p_{1:d}$ if and only if 
\begin{enumerate}
\item $p_k$ is conservative given $M_{0:k-1}$ and $U_{k-1}$, and \RTcomment{clearer to say ``sub-uniform''?}
\item $p_k$ is a function of $M_{0:k}$ and $U_k$.
\end{enumerate}

@@

\section{Discussion}
\label{sec:discussion}
